# The Billboard Hot 100 #1 Songs: What the Top Tells Us About the Evolution of Music

```{r}
library(tidyverse)
library(lubridate)
library(scales)
library(redav)
library(ggplot2)
library(ggrepel)
library(GGally)
library(vcd)
library(ggmosaic)
library(ggalluvial)
library(patchwork)

billboard <- read_csv("data/billboard_clean.csv", show_col_types = FALSE)
```

## 🎵 "You're Still The One" 🎵

Have you ever listened to a song so good you just had to play it over and over again... for weeks? Here, we ask ourselves what that might look like in the Billboard Hot 100. Since the list is updated weekly, it is totally possible for a song to show up across multiple weeks in the span of a year. The question is: how often does that occur every decade?

### Unique Songs that Hit #1 Across Decades

```{r}
#updating billboard to include a column for year and decade
billboard_updated <- billboard |>
  mutate(
    date_clean = as.Date(.data$date),  # Convert POSIXct to Date to help render
    year = lubridate::year(date_clean),  # Extract year
    decade = floor(year / 10) * 10
  ) |>
  select(-date) |>  # Remove old date column
  rename(date = date_clean)  # Rename back to date


week_level <- billboard_updated |>
  # Create a vector of Date objects, one for each week the song was #1.
  # map2() applies a function to (date, weeks_at_number_one) row-by-row.
  # seq.Date() generates weekly sequences like: date, date+7, date+14, ...
  mutate(week_dates = map2(
    .data$date, 
    weeks_at_number_one,
    ~ seq.Date(as_date(.x), by = "week", length.out = .y)
  )) |>
  # Expand the list-column `week_dates` so each row becomes one weekly observation.
  unnest(cols = c(week_dates)) |>
  rename(chart_date = week_dates) |>
  # Extract additional useful time components from the chart week
  mutate(
    year = year(chart_date),
    week_of_year = isoweek(chart_date),
    song_id = paste0(song, " — ", artist)
  ) |>
  arrange(chart_date)

# Summary: number of unique #1 songs per year
songs_per_year <- billboard_updated |>
  group_by(year) |>
  summarise(unique_no1 = n_distinct(paste0(song, " — ", artist)), .groups = "drop")
```

How many different songs hit #1 in 1975 vs. 2015?

```{r}
# Print numbers for 1975 vs 2015
val_1975 <- songs_per_year |> filter(year == 1975) %>% pull(unique_no1)
val_2015 <- songs_per_year |> filter(year == 2015) %>% pull(unique_no1)
cat("Unique #1 songs in 1975:", ifelse(length(val_1975)==0, NA, val_1975), "\n")
cat("Unique #1 songs in 2015:", ifelse(length(val_2015)==0, NA, val_2015), "\n")
```

From our initial exploration, we noticed that in 1975 there were 35 unique #1 songs, while in 2015 there were only 8. This suggests that in earlier decades, more songs reached the top of the chart briefly, whereas in recent years, fewer songs dominate the #1 spot for longer periods.

To investigate this further, we can examine the distribution of weeks_at_number_one across decades. By creating boxplots for each decade, we can visually compare: - The median number of weeks songs stayed at #1 in each decade. - How the range and spread of chart-topping durations changed over time. - Any extreme long-running hits (outliers) that may skew the averages in recent decades.

This approach helps confirm whether songs in recent decades truly stay at #1 longer than in earlier decades, supporting the hypothesis of increased chart dominance by fewer songs.

If songs today tend to dominate the charts for longer periods than in the past, then in the boxplot of weeks_at_number_one by decade, we would expect to see: - Earlier decades (e.g., 1970s, 1980s): Lower median and shorter interquartile range, indicating most songs spent only a few weeks at #1. - Recent decades (e.g., 2000s, 2010s): Higher median and longer upper whiskers, showing that fewer songs reached #1 but those that did stayed there for many weeks. - Outliers: More extreme long-running #1 hits in recent decades (longer upper tails), reflecting chart dominance by a small number of songs.

```{r}
# Plot 1: Box plot of weeks at #1 across decades
# Ensure each song has a single decade (counts the first decade it shows up in)
billboard_unique <- billboard_updated |>
  group_by(song) |>
  summarize(decade = first(decade), .groups = "drop")

# Prepare weeks_by_song + decade

weeks_by_song <- billboard_updated |>
  distinct(song, artist, weeks_at_number_one) |>
  mutate(weeks_at_number_one = as.integer(weeks_at_number_one))

weeks_with_decade <- weeks_by_song |>
  left_join(billboard_unique, by = "song") |>
  mutate(decade = factor(decade))  # ensure decade is categorical

# Boxplot of weeks at #1 by decade
p_box <- ggplot(weeks_with_decade, aes(x = decade, y = weeks_at_number_one)) +
  geom_boxplot(outlier.size = 1, varwidth = TRUE, fill = "skyblue") +
  labs(
    title = "Distribution of Weeks at #1 by Decade",
    x = "Decade",
    y = "Weeks at #1"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p_box)
```

-   Median weeks at #1 remains consistent at \~2 weeks across all early decades
-   Medians increase slightly in 2000s-2010s, suggesting songs stay at #1 longer on average
-   Gradual increase in extreme outliers from 1950s through 1990s (blockbuster hits becoming more common)
-   2020 already shows 7 outliers despite partial data, suggesting most extreme chart concentration of any era despite only having only 5 years of data

Consistent medians (1950s-1980s): The median (thick black line) stays remarkably stable around 2 weeks at #1 across the first several decades. This means that in any of these decades, the typical #1 song held the position for about 2 weeks.

Shift in recent decades (2000s-2010s): The medians appear to increase slightly, and the boxes themselves shift upward. This indicates that songs reaching #1 in recent decades tend to stay there longer on average.

Outliers: From the 1950s through 1990s, we see a gradual increase in extreme long-running hits, suggesting that blockbuster songs became slightly more common. However, the dramatic spike to 7 outliers in the 2020s is striking and suggests a fundamental shift in how the charts work in the streaming era.

While it's true that fewer songs dominate the charts in recent years, the story is more about extremes than overall shifts. Most songs still cluster around 2-3 weeks at #1 even in recent decades, but there's an increasing tendency for certain mega-hits to stay at #1 for extraordinarily long periods (10+ weeks or even 15+ weeks based on those outliers). This creates a "winner-take-all" dynamic where streaming platforms and algorithmic recommendations may help a few songs achieve unprecedented chart dominance while the typical song's chart tenure remains similar to earlier eras.

Keeping in mind that 2020s is incomplete becuase we are still in the decade. The 2020s data is essentially showing a snapshot in progress rather than a complete picture. The fact that we already have 7 outliers in just 5 years suggests that the full decade would likely show the most extreme chart concentration of any era, with both a higher typical duration AND more mega-hits dominating for months at a time. Even with incomplete data, the 2020s are already showing unprecedented chart dominance by a small number of songs.

### Marathon Hits (Long Reigning #1s) Across Decades

To better understand these outliers, I wanted to identify the top 10 longest-reigning #1 songs of all time. Additionally, I was curious to examine the longest-running #1 hit from each decade to see if any songs maintained their dominance across multiple decades or if each era had its own distinct chart champions.

```{r}
# Identify marathon hits (10+ weeks)
song_runs <- billboard_updated |>
  distinct(song, artist, weeks_at_number_one, year, decade) |>
  mutate(
    song_id = paste0(song, " — ", artist),
    weeks_at_number_one = as.integer(weeks_at_number_one)
  )

top10_alltime <- song_runs |>
  arrange(desc(weeks_at_number_one)) |>
  slice_head(n = 10)

longest_per_decade <- song_runs |>
  group_by(decade) |>
  slice_max(weeks_at_number_one, n = 1, with_ties = FALSE) |>
  ungroup()

marathon_hits <- song_runs |>
  mutate(type = ifelse(weeks_at_number_one >= 10, "marathon", "typical"))
```

```{r}
top10_alltime
```

```{r}
longest_per_decade
```

The data reveals a striking pattern of increasing chart dominance over time. While the longest-reigning #1 songs from the 1950s through 1980s peaked at just 9-10 weeks, this ceiling shattered in the 1990s when Mariah Carey's "One Sweet Day" held the top spot for 16 weeks. The trend accelerated dramatically in the streaming era: 7 of the top 10 longest-reigning songs of all time are from 2017 or later, with the 2020s claiming the record at 19 weeks despite being an incomplete decade. Notably, the 2020s already has three songs in the all-time top 10—more than any complete decade from the 20th century. This confirms the "winner-take-all" dynamic identified earlier, where streaming platforms and algorithmic recommendations enable a small number of mega-hits to achieve unprecedented chart dominance, fundamentally transforming how songs reach and maintain the #1 position.

Next, we wanted to investigate whether there are seasonal patterns in when marathon hits (songs that topped the charts for 10+ weeks) first reached #1. Specifically, we were curious if these dominant songs tend to emerge during particular times of the year, such as summer months when listening increases or the December holiday season when certain songs gain cultural momentum.

```{r}
# Plot 2: Bar Chart of Marathon Hits by Month
# Count marathon hits per month
marathon_month_counts <- marathon_hits_only |>
  count(month)

# Bar chart of marathon hits by month
ggplot(marathon_month_counts, aes(x = month, y = n)) +
  geom_col(fill = "steelblue", color = "black", width = 0.9) +
  labs(
    title = "Month When Marathon Hits First Reached #1",
    subtitle = "Songs with 10+ consecutive weeks at #1",
    x = "Month",
    y = "Number of Marathon Hits"
  ) +
  theme_minimal(base_size = 14)
```

-   April and November lead with 6 marathon hits each—significantly above other months
-   November's success likely driven by holiday season momentum carrying songs into the new year
-   April's peak lacks obvious cultural explanation, making it the most surprising finding
-   December produced only 3 marathon hits despite being peak holiday month
-   No single "magic month" guarantees marathon success—factors are more complex than seasonal listening alone

The bar chart reveals interesting seasonal patterns in when marathon hits (songs with 10+ consecutive weeks at #1) first reach the top of the charts. April and November emerge as the most successful months, each producing 6 marathon hits—significantly more than any other time of year. This bimodal distribution suggests that timing does play a role in a song's potential for extended chart dominance, though the underlying factors differ between these two peaks.

The November spike is particularly intriguing and can likely be attributed to strategic timing around the holiday season. Songs that reach #1 in November benefit from sustained listening through the holidays and into the new year, giving them momentum to maintain their position for months. However, December itself only produced 3 marathon hits, suggesting that releases during the actual holiday season face fiercer competition or stronger turnover as listeners shift to new music in January.

Surprisingly, the data challenges the assumption that summer is prime time for chart-dominating hits. While June shows moderate success with 5 marathon hits, July and August produced only 3-4 each. This mixed performance suggests there's no clear "summer song" advantage for achieving prolonged dominance, despite increased listening during warmer months.

The most puzzling finding is the April peak, which produced 6 marathon hits despite lacking an obvious cultural explanation like holidays or seasonal listening patterns. Meanwhile, February, March, and September emerge as "dead zones" with only 2 marathon hits each—the lowest counts on the chart. Overall, while the data shows that timing matters, there's no single "magic month" that guarantees a marathon hit, and the success factors appear more complex than simple seasonal listening patterns.

### One Week Wonders Across Decades

While looking and investigating marathons which is a smalle proportion of the data thought we would be able to learn more from looking into one week wonder where a song charted only for one week. Wanted to first look what percentage of the songs only charted for one week.

```{r}
# Calculate percentage of #1 songs with only 1 week at the top
one_week_pct <- weeks_by_song |>
  summarise(
    one_week_count = sum(weeks_at_number_one == 1, na.rm = TRUE),
    total_songs = n(),
    pct = one_week_count / total_songs * 100
  )

one_week_pct
```

That means roughly 1 in 3 #1 hits were “one-week wonders,” topping the chart briefly before being replaced.

```{r}
# Compute summary per decade
one_week_by_decade <- billboard_updated |>
  distinct(song, artist, decade, weeks_at_number_one) |>   # one row per song
  group_by(decade) |>
  summarise(
    total_songs = n(),
    one_week_count = sum(weeks_at_number_one == 1),
    pct_one_week = one_week_count / total_songs * 100
  ) |>
  arrange(decade)

one_week_by_decade
```

Looking across decades, the share of songs that spend only a single week at #1 reveals a striking pattern in chart behavior. In the 1950s and 1960s, one-week #1 hits were relatively rare, typically making up around a quarter or less of all chart-toppers. But in the 1970s this changed dramatically—nearly 50% of #1 songs that decade lasted only a single week, indicating much higher turnover at the top of the chart.

This proportion declines again in the 1980s through the 2010s, suggesting that songs generally enjoyed longer runs at #1 during those decades. However, the 2020s reverse that trend sharply: over 56% of #1 songs so far have lasted just one week. This aligns with an era defined by rapid consumption cycles, algorithmic feeds, and social-media–driven virality, where hits can spike quickly and disappear just as fast.

Interestingly, this pattern complicates the earlier hypothesis that songs now tend to dominate the charts for longer. Both things appear to be true: the longest #1 runs are getting longer (the rise of marathon hits), but simultaneously more songs are lasting only one week. In other words, chart behavior is becoming increasingly polarized—a mix of very short-lived hits and a few extremely dominant ones.

To better understand these one-week wonders, we investigated their genres across the entire dataset. By examining the types of songs that only spent a single week at #1, we can identify whether certain musical styles are more prone to brief chart dominance. This helps reveal patterns in listener behavior, production trends, and how certain genres may be more likely to generate short-term viral hits versus sustained popularity.

?? Change this to answer a different question doesn't seem like an interesting result

```{r}
# Filter only one-week #1 hits
one_week_hits <- billboard_updated %>%
  filter(weeks_at_number_one == 1)

# Count number of one-week hits per cdr_genre
genre_counts <- one_week_hits %>%
  group_by(cdr_genre) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

# Convert to factor for ordered bars
genre_counts$cdr_genre <- factor(genre_counts$cdr_genre, levels = genre_counts$cdr_genre)

# Bar chart
ggplot(genre_counts, aes(x = cdr_genre, y = count)) +
  geom_col(fill = "steelblue", color = "black", width = 0.8) +
  labs(
    title = "Genres of One-Week #1 Hits",
    subtitle = "Songs that spent only a single week at the top of the charts",
    x = "Genre",
    y = "Number of One-Week Hits"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

------------------------------------------------------------------------

## 🎵 "I've Been Afraid of Changing" 🎵

The world looks different now in 2025 than it did in 1950, and so did the music we listen to. Here, we explore how music has changed across decades. Do they sound different musically? Are we using different instruments than we did 70 years ago? Are we listening to more singer-songwriters than bands?

### Instrumentation and Sound Evolution

```{r}
song_vars <- billboard_updated |>
  select(song, decade, bpm, energy, danceability, happiness, loudness_d_b,
         acousticness, length_sec, instrumental_length_sec)
song_vars <- song_vars |> drop_na()

pca_all <- prcomp(song_vars[, 3:ncol(song_vars)], scale. = TRUE)
round(pca_all$rotation, 5)
pc_scores_all <- pca_all$x
rownames(pc_scores_all) <- song_vars$song 
pca_summary <- summary(pca_all)
round(rbind(pc_scores_all, NA, pca_summary$importance), 5)
```

```{r}
billboard_20th <- song_vars |> filter(decade < 2000)
billboard_21st <- song_vars |> filter(decade >= 2000)

song_vars_20th <- billboard_20th |> 
  select(bpm, energy, danceability, happiness, loudness_d_b,
         acousticness, length_sec, instrumental_length_sec)

song_vars_21st <- billboard_21st |> 
  select(bpm, energy, danceability, happiness, loudness_d_b,
         acousticness, length_sec, instrumental_length_sec)
```

```{r}
# Plot 3
draw_biplot(song_vars_20th, label_size = 2) +
  ggtitle("Biplot: Songs Before 2000")
```

```{r}
# Plot 3
draw_biplot(song_vars_21st, label_size = 3) +
  ggtitle("Biplot: Songs From 2000 Onward")
```

?? How do I remove the bullets? Should I keep? I am just interpreting the vectors

The relationships between musical features have changed, suggesting that modern songs combine these characteristics differently than older songs.

-   Before 2000, danceability, energy, and loudness were strongly related.
-   Post-2000, danceability and acousticness appear uncorrelated, unlike before.
-   Before 2000, BPM and happiness were positively correlated; after 2000, they no longer align.

```{r}
# Convert decade to a factor
song_vars$decade <- as.factor(song_vars$decade)
# Plot 5
# Parallel coordinate plot
ggparcoord(
  data = song_vars,
  columns = 3:ncol(song_vars),  # numeric variables
  groupColumn = "decade",       # color lines by decade
  scale = "std",          # scale variables 0-1
  showPoints = FALSE,           # optionally show points
  alphaLines = 0.5              # transparency
) +
  scale_color_brewer(palette = "Set1") +
  theme_minimal() +
  labs(title = "Parallel Coordinate Plot of Billboard Songs by Decade",
       x = "Song Attributes",
       y = "Scaled Value")

```

?? I know she said not to make any interactive plots but it would also be helpful to see the relationships between the variables. Maybe I should just do it for myself and place the variables next to each other in an order that I can make observations

When I use the std scale it reveals which decades are above/below average for each attribute.

-   1960s songs have shorter length_sec.
-   1970s songs show higher instrumental_length_sec.
-   2000s–2020s move similarly in loudness vs. acousticness; 1960s–1970s show increased acousticness.
-   Energy, Danceability, Happiness cluster around 0 across decades, indicating consistency over time.

### Evolution of Genre Across Decades

When I think 1960s-70s, I think of rock as the defining music genre of that time. When I think 2010-2020s, I think of indie, hip-hop, and pop. But will this reflect in the Billboard Hot 100? To test my hypothesis, I create a heatmap of the proportion of genres of #1 songs for each decade. Since songs could have more than one genre, I define the proportion of a genre as how many times a genre appeared in that decade divided by the total number of songs that appeared in that decade.

```{r}
# Genre representation

# Separate rows with multiple genres
genres_billboard <- billboard_updated |>
  separate_rows(discogs_genre, sep=";") |>
  mutate(genre_sep = str_trim(discogs_genre)) |>
  filter(genre_sep != "")

# Count songs by decade and genre
genre_counts <- genres_billboard |>
  group_by(decade, genre_sep) |>
  summarise(n_genre = n(), .groups = "drop")

# Count unique songs per decade
decade_totals <- billboard_updated |>
  group_by(decade) |>
  summarise(n_total = n(), .groups = "drop")

# Calculate proportion of genre per decade
genre_prop <- genre_counts |>
  left_join(decade_totals, by = "decade") |>
  mutate(proportion = n_genre/n_total)

# Heatmap data
heatmap_genre <- genre_prop |>
  select(decade, genre_sep, proportion)

# Ordering by average proportion for genres
genre_avg <- heatmap_genre |>
  group_by(genre_sep) |>
  summarise(avg_prop = mean(proportion, na.rm = TRUE)) |>
  arrange(avg_prop)

heatmap_genre <- heatmap_genre |>
  mutate(genre_sep = factor(genre_sep, levels = genre_avg$genre_sep))

# Create the heatmap
ggplot(heatmap_genre, aes(x = decade, y = genre_sep, fill = proportion)) +
  geom_tile() +
  scale_fill_distiller(palette = "RdPu", direction = 1, name = "Genre Proportion") +
  # scale_fill_viridis_c(name = "Genre Proportion", direction = -1) +
  scale_x_continuous(breaks=unique(heatmap_genre$decade)) +
  labs(
    title = "Heatmap of Genre per Decade",
    subtitle = "In decreasing order of average proportion",
    x = "Decade",
    y = "Genre") +
  theme_minimal()
```

The heat map above shows us that Pop has always been a popular (haha) genre, which is not surprising. It lost to Rock as the genre with the most #1 songs in the 60s and 70s, though. We also see that Hip Hop as a genre did not even show up in any of the #1 songs until the 1980s, and since then has consistently charted, with the peak being in the 2000s. Almost opposite to Hip Hop is Rock, which was prominent in the 50s to 80s, but quickly lost its popularity from the 1990s onwards.

### Solo/Duo/Group Artists vs Featured Collaborations

Just like I picture the 60s and 70s to be the defining decades for Rock, I also picture the 60s and 70s to be the defining decades for bands. Names like "The Beatles" and "Fleetwood Mac" - both bands - come to mind as prominent artists of that era, whereas names like "Taylor Swift" and "Ariana Grande" - solo artists - come to mind as prominent artists of the era we are in right now. To test this hypothesis, I create a mosaic plot that shows us the distribution of artist structure (group, duo, solo) and featured collaborations per decade.

```{r}
# Solo artists vs groups vs collaborations

# Separating between solo, duo and group
# Separating between feature vs no feature
bb_artist <- billboard_updated |>
  mutate(
    artist_type = case_when(
      artist_structure %in% c(0, 0.5) ~ "Group",
      artist_structure %in% c(1, 1.5) ~ "Solo",
      artist_structure %in% c(2, 2.5) ~ "Duo"
    ),
    has_feature = ifelse(artist_structure %% 1 == 0.5, "With Feature", "No Feature"),
    artist_type = factor(artist_type, levels = c("Group", "Duo", "Solo")),
    era_detailed = case_when(
      decade < 1990 ~ "Before 1990", 
      decade >= 1990 & decade < 2010 ~ "1990-2009",
      decade >= 2010 ~ "2010-2025"
      ),
    era_detailed = factor(era_detailed, levels = c("Before 1990", "1990-2009", "2010-2025"))
  )

# Stacked bar chart with artist structure
structure_chart <- ggplot(data = bb_artist, aes(x = decade, fill = artist_type)) +
  geom_bar(position = "fill") +
  scale_x_continuous(breaks = unique(bb_artist$decade), trans = "reverse") +
  scale_fill_manual(
    values = c("Solo" = "lightsteelblue2", "Duo" = "lightsteelblue3", "Group" = "lightsteelblue4"),
    name = "Artist Structure"
  ) +
  labs(
    title = "How Has Artist Structure Changed Across Decades?",
    x = "Decade",
    y = "Song Proportion"
  ) +
  coord_flip()


# Stacked bar chart with artist feature
feature_chart <- ggplot(data = bb_artist, aes(x = era_detailed, fill = has_feature)) +
  geom_bar(position = "fill") +
  scale_fill_manual(
    values = c("With Feature" = "red4", "No Feature" = "lightgoldenrod"),
    name = "Feature"
  ) +
  labs(
    title = "How Have Music Collaborations Changed?",
    x = "Era",
    y = "Proportion of Features"
  )

structure_chart + feature_chart
```

These two stacked bar charts confirm my hypothesis. Firstly, it shows us that group artists are slowly becoming less popular every decade. In the 1960s, group artists made up most of the #1 songs, where now, it is dominated by solo artists. Secondly, it shows us that virtually no featured collaborations made it to #1 before the 1990s, which is consistent with a [2018 report from The Economist](https://www.economist.com/graphic-detail/2018/02/02/popular-music-is-more-collaborative-than-ever) telling us that collaboration in music is a relatively recent phenomena.

------------------------------------------------------------------------

## 🎵 "Say My Name, Say My Name" 🎵

The songs we listen to might sound different now, but could it be because of the people behind the curtain? Here, we pay homage to the artists, songwriters, and producers working hard behind the scenes to give us the songs that eventually show up as #1 in the Billboard Hot 100. How have they changed across decades?

### Top Artists Over the Years

Our data spans 1950 to 2025, so I am interested in looking at the top artists of all time (artists that have the most songs at #1 in the Billboard Hot 100) and what years they were popular. For instance, were The Beatles only popular for a certain time or do they have songs that are #1 even now?

```{r}
# Get top 20 artists
top_20_artists <- billboard_updated |>
  count(artist, sort = TRUE) |>
  slice_head(n = 20) |>
  pull(artist)

# Get first and last year for each artist
artist_timeline <- billboard_updated |>
  filter(artist %in% top_20_artists) |>
  group_by(artist) |>
  summarise(
    first_year = min(year, na.rm = TRUE),
    last_year = max(year, na.rm = TRUE),
    n_songs = n()
  ) |>
  arrange(first_year, desc(n_songs)) |>
  mutate(artist = factor(artist, levels = rev(artist)))

# Get individual songs for points
artist_songs <- billboard_updated |>
  filter(artist %in% top_20_artists) |>
  select(artist, year) |>
  left_join(artist_timeline |> select(artist), by = "artist") |>
  mutate(artist = factor(artist, levels = levels(artist_timeline$artist)))

# Create timeline plot
ggplot(artist_timeline, aes(y = artist)) +
  geom_segment(aes(x = first_year, xend = last_year, yend = artist),
               alpha = 1, size = 0.5, color = "black") +
  geom_point(data = artist_songs, aes(x = year, y = artist), 
             color = "maroon", alpha = 0.5, size = 4) +
  labs(
    title = "Active Years for Top 20 Artists with #1 Songs",
    subtitle = "Each dot represents a song that charted as #1",
    x = "Year",
    y = "Artist"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )
```

The chart above takes the top 20 artists of all time and marks the first year and the most recent year in which their song hit #1. While most of these artists are not popular for more than 10 years, there are some interesting anomalies. Mariah Carey, for example, first hit #1 in the mid 1990s and last hit #1 in 2019, although it looks she was more consistently popular in the 1990s, falling off after 2000. The Beatles and The Supremes have almost the opposite story, with their reign of popularity only taking up a few years despite having many more songs hit #1.

### Are #1 Artists & Songwriters Becoming More Diverse?

With more and more opportunities being presented to artists of color and women artists over the years, I would imagine there being a stark difference between the gender and race distribution of #1 artists now than in 1950. I test this hypothesis by creating two mosaic plots exploring the effect of gender and race on era (defined as before 1990 and after 1990).

```{r}
# Artist Gender and Race
artist_data <- billboard_updated |>
  mutate(
    gender = case_when(
      artist_male == 0 ~ "All Female",
      artist_male == 1 ~ "All Male",
      artist_male == 2 | artist_male == 3 ~ "Mixed Gender"
    ),
    race = case_when(
      artist_white == 0 ~ "Not White",
      artist_white == 1 ~ "White",
      artist_white == 2 ~ "Mixed Race"
    ),
    era_detailed = case_when(
      decade < 1990 ~ "Before 1990", 
      decade >= 1990 ~ "After 1990"
      ),
    era_detailed = factor(era_detailed, levels = c("Before 1990", "After 1990")),
    gender = factor(gender, levels = c("All Male", "All Female", "Mixed Gender"))
  ) |>
  select(artist, decade, gender, race, era_detailed) |>
  distinct()

# define grayscale palette
shades <- c("Mixed Race" = gray(0.8), 
            "Not White" = gray(0.5), 
            "White" = gray(0.2))

vcd::mosaic(~era_detailed + gender + race,
       data = artist_data,
       direction = c("v", "v", "h"),
       main = "Artist Gender and Race Interaction",
       highlighting = "race",
       highlighting_fill = shades,
       main_gp = gpar(fontsize = 12),
       labeling_args = list(
         gp_labels = gpar(fontsize = 5),   # Adjust category label font size
         gp_varnames = gpar(fontsize = 7) # Adjust variable names font size
       ))
```

The mosaic plot above confirms my hypothesis and shows us that, before 1990, "All Male" seemed to take up the biggest proportion of artists, with more than 2/3 of these male artists being white. "All Female" and "Mixed Gender" artists had similar proportions, but both were still overwhelmingly white.

After 1990, it still looks like "All Male" artists dominate #1 compared to "All Female" and "Mixed Gender" artists, but the distribution of race makes a dramatic 180. Now, half of the male artists are not white, compared to before 1990 where not white artists only made up less than 1/3 of all male artists. This is a huge improvement for the distribution of race in our #1 artists!

```{r}
# Top Songwriters
songwriter_data <- billboard_updated |>
  mutate(
    gender = case_when(
      songwriter_male == 0 ~ "All Female",
      songwriter_male == 1 ~ "All Male",
      songwriter_male == 2 | songwriter_male == 3 ~ "Mixed Gender"
    ),
    race = case_when(
      songwriter_white == 0 ~ "Not White",
      songwriter_white == 1 ~ "White",
      songwriter_white == 2 ~ "Mixed Race"
    ),
    era_detailed = case_when(
      decade < 1990 ~ "Before 1990", 
      decade >= 1990 ~ "After 1990"
      ),
    era_detailed = factor(era_detailed, levels = c("Before 1990", "After 1990")),
    gender = factor(gender, levels = c("All Male", "All Female", "Mixed Gender"))
  ) |>
  select(songwriters, gender, race, era_detailed) |>
  drop_na() |>
  distinct()

# define grayscale palette
shades <- c("Mixed Race" = gray(0.8), 
            "Not White" = gray(0.5), 
            "White" = gray(0.2))

vcd::mosaic(~era_detailed + gender + race,
       data = songwriter_data,
       direction = c("v", "v", "h"),
       main = "Songwriter Gender and Race Interaction",
       highlighting = "race",
       highlighting_fill = shades,
       main_gp = gpar(fontsize = 12),
       labeling_args = list(
         gp_labels = gpar(fontsize = 5),   # Adjust category label font size
         gp_varnames = gpar(fontsize = 7) # Adjust variable names font size
       ))
```

For songwriters, it looks like before 1990, it was dominated by songwriting groups of all white males.

After 1990, however, we once again see a big improvement. We now have a more even split between songwriting groups that are all male and mixed gender (all female groups are extremely rare, apparently). Within both male and mixed gender groups, we see a more even split between mixed race, not white, and white songwriters. Interestingly, both all male and mixed gender groups are dominated by mixed race songwriters!

From the two mosaic plots above, we can conclude that we are indeed seeing a big jump in diversity in the artists and songwriters behind our #1 songs in the Billboard Hot 100. A win for the music industry!
